A \proglang{Python}/\proglang{Zig} optimized and customizable implementation for the $\rho_{DCCA}$ and $DMC_x^2$ methods

### 2024-11-01 paper_v01/article.tex
DMCx2 flowchart update

%% - \Abstract{} almost as usual
\Abstract{
  This paper presents tha \pkg{Zebende}, a \proglang{Python} package writen in \proglang{Python} and \proglang{Zig}, that calculates the $DFA$, $DCCA$ $\rho_{DCCA}$ and the $DMC_x^2$. The package presents an optimized algorithm that significantly improves the calculations speed. A comparison with ohter packages that calculates the .The package is also the firts to implement the $DMC_x^2$ corefficient for any number of series.
}

@@

@@

@@

@@

@@

\begin{equation}
  \rho_{Y,X_k}(n)^T=[\rho_{Y,X_1}(n), \rho_{Y,X_2}(n),\cdots,\rho_{Y,X_j}(n)]
  \label{eq:rhocol}
\end{equation}

\begin{description}

  \label{steps:dfa}
  \item Taking a time series \(\{x_{i}\}\) with \(i\) ranging from \(1\) to \(N\), the integrated series \(X_{k}\) is calculated by \(X_{k} = \sum_{i=1}^{k}\left[x_{i} - \langle x \rangle \right] \) with \(k\) also ranging from \(1\) to \(N\);
  \item the  \(X_{k}\) series is divided in \(N - n\) boxes of size \(n\)(time scale),each box containing \(n + 1\) values, starting in \(i\) up to \(i + n\);
  \item for each box, a polynomial (usually of degree 1) best fit is calculated, getting \(\widetilde{X}_{k, i}\) with \( i \le k \le (i + n) \) (detrended values);
  \item in each box is calculated: \(f_{DFA}^{2}(n, i) = \frac{1}{1+n} \sum_{k=i}^{i + n}(X_{k}-\widetilde{X}_{k, i})^{2}\)
  \item for all the boxes of a time scale, the DFA is calculated as:\\[10pt]
        \(F_{DFA}(n) = \sqrt{\frac{1}{N - n} \sum_{i=1}^{N-n} f_{DFA}^{2}(n, i)}\);
  \item for a number of different timescales (n), with possible values \( 4 \le n \le \frac{N}{4}\) the \(F_{DFA}\) function is calculated to find a relation among \(F_{DFA} \times n\)

\end{description}

\begin{description}
  \label{steps:DCCA}
  \item Taking two time series with the same length \(\{x\alpha_{i}\}\) and \(\{x\beta_{i}\}\) with \(i\) ranging from \(1\) to \(N\),
        the integrated series \(X\alpha_{k}\) and \(X\beta_{k}\) are calculated by
        \(X_{k} = \sum_{i=1}^{k}\left[x_{i} - \langle x \rangle \right] \) for each series, with \(k\) also ranging from \(1\) to \(N\);
  \item \(X\alpha_{k}\) and \(X\beta_{k}\) series are divided in \(N - n\) boxes of size \(n\)(time scale),
        each box containing \(n + 1\) values, starting in \(i\) up to \(i + n\);
  \item for each box, a polynomial (usually of degree 1) best fit is calculated, getting
        \(\widetilde{X\alpha}_{k, i}\) and \(\widetilde{X\beta}_{k, i}\),
        for series \(\{x\alpha_{i}\}\) and \(\{x\beta_{i}\}\) respectively,
        with \( i \le k \le (i + n) \) (detrended values);
  \item in each box is calculated: \(f_{DCCA}^{2}(n, i) =
        \frac{1}{1+n} \sum_{k=i}^{i + n}(X\alpha_{k}-\widetilde{X\alpha}_{k, i}) \times (X\beta_{k}-\widetilde{X\beta}_{k, i})\)
  \item for all the boxes of a time scale, the DFA is calculated as:\\[10pt]
        \(F_{DCCA}^{2}(n) =\frac{1}{N - n} \sum_{i=1}^{N-n} f_{DCCA}^{2}(n, i)\);
  \item for a number of different timescales (n), with possible values
        \( 4 \le n \le \frac{N}{4}\) the \(F_{DCCA}^{2}\) function is calculated to find a relation among
        \(F_{DCCA}^{2} \times n\)
\end{description}

@@

\begin{leftbar}
  As usual \dots
\end{leftbar}

>>>>

%% - \Abstract{} almost as usual
\Abstract{
  This paper presents tha \pkg{Zebende}, a \proglang{Python} package written in \proglang{Python} and \proglang{Zig}, that calculates the $DFA$, $DCCA$ $\rho_{DCCA}$ and the $DMC_x^2$. The package presents an optimized algorithm that significantly improves the calculations speed. A comparison with other packages that calculates the .The package is also the first to implement the $DMC_x^2$ coefficient for any number of series.
}

@@

% rho DCCA applications

@@

% DMCx2 applications

@@

The algorithms that calculates the $\rho_{DCCA}$ uses the $DFA$ and the $DCCA$ steps. The $DMC_x^2$ coefficient uses the $\rho_{DCCA}$ coefficient and, consequently, also embraces the $DFA$ and the $DCCA$. The $DFA$ method is described in six steps:

\begin{enumerate}\label{steps:dfa}

  \item Taking a time series \(\{x_{i}\}\) with \(i\) ranging from \(1\) to \(N\), the integrated series \(X_{k}\) is calculated by \(X_{k} = \sum_{i=1}^{k}\left[x_{i} - \langle x \rangle \right] \) with \(k\) also ranging from \(1\) to \(N\);
  \item the  \(X_{k}\) series is divided in \(N - n\) boxes of size \(n\)(time scale), each box containing \(n + 1\) values, starting in \(i\) up to \(i + n\);
  \item for each box, a polynomial (usually of degree 1) best fit is calculated, getting \(\widetilde{X}_{k, i}\) with \( i \le k \le (i + n) \) (detrended values);
  \item in each box is calculated: \(f_{DFA}^{2}(n, i) = \frac{1}{1+n} \sum_{k=i}^{i + n}(X_{k}-\widetilde{X}_{k, i})^{2}\)
  \item for all the boxes of a time scale, the $DFA$ is calculated as:\\[10pt]
        \(F_{DFA}(n) = \sqrt{\frac{1}{N - n} \sum_{i=1}^{N-n} f_{DFA}^{2}(n, i)}\);
  \item for a number of different timescales (n), with possible values \( 4 \le n \le \frac{N}{4}\) the \(F_{DFA}\) function is calculated to find a relation among \(F_{DFA} \times n\)

\end{enumerate}

The $DCCA$ method is very similar to the $DFA$ calculations. With the difference of analyzing two series while the $DFA$ evaluate properties of a single time series. It´s also a six steps process:

\begin{enumerate}
  \label{steps:DCCA}
  \item Taking two time series with the same length $\{x\alpha_{i}\}$ and \(\{x\beta_{i}\}\) with \(i\) ranging from \(1\) to \(N\),
        the integrated series \(X\alpha_{k}\) and \(X\beta_{k}\) are calculated by
        \(X_{k} = \sum_{i=1}^{k}\left[x_{i} - \langle x \rangle \right] \) for each series, with \(k\) also ranging from \(1\) to \(N\);
  \item \(X\alpha_{k}\) and \(X\beta_{k}\) series are divided in \(N - n\) boxes of size \(n\)(time scale), each box containing \(n + 1\) values, starting in \(i\) up to \(i + n\);
  \item for each box, a polynomial (usually of degree 1) best fit is calculated, getting
        $\widetilde{X\alpha}_{k, i}$ and $\widetilde{X\beta}_{k, i}$,
        for series $\{x\alpha_{i}\}$ and $\{x\beta_{i}\}$ respectively,
        with \( i \le k \le (i + n) \) (detrended values);
  \item in each box is calculated: $f_{DCCA}^{2}(n, i) =
        \frac{1}{1+n} \sum_{k=i}^{i + n}(X\alpha_{k}-\widetilde{X\alpha}_{k, i}) \times (X\beta_{k}-\widetilde{X\beta}_{k, i})$
  \item for all the boxes of a time scale, the $DCCA$ is calculated as:\\[10pt]
        $F_{DCCA}^{2}(n) =\frac{1}{N - n} \sum_{i=1}^{N-n} f_{DCCA}^{2}(n, i)$;
  \item for a number of different timescales (n), with possible values
        $4 \le n \le \frac{N}{4}$ the $F_{DCCA}^{2}$ function is calculated to find a relation among
        $F_{DCCA}^{2} \times n$
\end{enumerate}

Comparing the algorithms, the first three are basically identical, the only difference is that the $DCCA$ method apply those steps to two series. The step four of the $DFA$ is essentially an application of the variance calculation and the equivalent step of the $DCCA$ is a covariance between the two series. Step five calculates the square root of the mean of the variances calculated in  in each box for the $DFA$, in the $DCCA$, the mean of the covariances calculated for each box is calculated in stead. The last step,in both cases, is more a reminder to repeat the respective previous operations for a number of difference time scales.

The $\rho_{DCCA}$ is measured using Eq. \ref{eq:p_dcca}. Considering the relation between $DFA$ and variance and $DCCA$ and covariance, the $\rho_{DCCA}$ resembles Pearson correlation for a time scale $n$.

The $DMC_x^2$ is a generalization of the $\rho_{DCCA}$ that calculates the correlation between one time-series $\left\lbrace Y \right\rbrace $, as the  dependent variable, and a number $j$ of time-series $\left\lbrace X_{1} \right\rbrace $, $\left\lbrace X_{2} \right\rbrace $, $\left\lbrace X_{3} \right\rbrace $, $\dots $, $\left\lbrace X_{j} \right\rbrace $ defined as independent variables. The coefficient is expressed mathematically as:

@@

In Eq. \ref{eq:dmcx2}, $\rho^{-1}(n)$ represent the inverse of a matrix populated by all possible combinations of $\rho_{DCCA}$ between independent variables. In Eq. \ref{eq:p_dcca_matrix}, value $\rho_{X_{1},X_{2}}(n)$, for instance, is the $\rho_{DCCA}$ for independent variables $X_{1}$ and $X_{2}$ calculated with time scale $n$, occupying position $\rho_{1 2}$ of the matrix. Two fundamental characteristics: the first is that the main diagonal values are all ones, since it's position in the matrix denotes the calculation of a correlation between a series and itself. Second, the matrix is symmetric in relation to the main diagonal, as the $\rho_{DCCA}$ is evaluated a commutative expression.

@@

At last Eq. \ref{eq:rhocol} represent the transposed vector of the $\rho_{Y,X_{i}}(n)$ between the depended variable $\left\lbrace Y \right\rbrace $ and each $\left\lbrace X_{i} \right\rbrace $ independent variable for a given time scale $n$.

\begin{equation} \label{eq:rhocol}
  \rho_{Y,X_i}(n)^T=[\rho_{Y,X_1}(n), \rho_{Y,X_2}(n),\cdots,\rho_{Y,X_j}(n)]
\end{equation}

The $\rho_{DCCA}$ and the $DMC_x^2$ should be evaluated in a number of time scales ($n$) to analyze the characteristics of each coefficient.

@@
